---
title: "Project 1"
author: "Daria Smyslova"
format: html
editor: visual
---

### Part 1: Data Processing

#### Parsing Variable Values

```{r}
library(httr)
library(jsonlite)

get_variable_values <- function(year) {
  if (length(year) != 1 || !year %in% 2021:2024) {
    stop("year must be 2021, 2022, 2023, or 2024")
  }

  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )

  response <- GET(url)
  stop_for_status(response)

  json_text <- content(response, as = "text", encoding = "UTF-8")
  variables <- fromJSON(json_text, simplifyVector = FALSE)$variables

  state_variable <- if (year <= 2022) "ST" else "STATE"

  keep <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_variable, "REGION", "DIVISION"
  )

  missing_variables <- setdiff(keep, names(variables))
  if (length(missing_variables) > 0) {
    stop(
      "Variables missing from the ", year, " JSON: ",
      paste(missing_variables, collapse = ", ")
    )
  }

  #Keep each requested variable's value definitions.
  lapply(variables[keep], function(variable) variable$values)
}
```

```{r}
#Run it once per year and save the four lists together:
values_2021 <- get_variable_values(2021)
values_2022 <- get_variable_values(2022)
values_2023 <- get_variable_values(2023)
values_2024 <- get_variable_values(2024)

all_variable_values <- list(
  `2021` = values_2021,
  `2022` = values_2022,
  `2023` = values_2023,
  `2024` = values_2024
)

dir.create("data", showWarnings = FALSE)
saveRDS(all_variable_values, file = "data/variable_values.rds")
```

Before querying the Census data, I downloaded the PUMS variable descriptions for 2021 through 2024. I wrote a function that sends a `GET()` request to the appropriate year’s JSON file, parses the response, and keeps only the variables needed for this project. The function accounts for a naming change in the state variable: it is called `ST` in 2021 and 2022 and `STATE` in 2023 and 2024. I ran the function for each year and saved the four resulting lists together in `data/variable_values.rds` so the later functions can use the correct codes and labels for each year.

```{r}
names(values_2021)             
names(values_2024)             
values_2024$SEX$item           
values_2024$AGEP$range         
readRDS("data/variable_values.rds")$`2024`$SEX$item
```

I checked the saved lists before moving on. The 2021 list contains `ST`, the 2024 list contains `STATE`, and the 2024 `SEX` values map codes `1` and `2` to “Male” and “Female.” I also inspected the `AGEP` range and read the `.rds` file back into R to confirm that it saved correctly.

#### Parsing API Data
